Convergence Studies and Effective Alphas for Disks with Gravitational Instabilities. \\ I. The Constant Cooling Case

### 2010-11-01 Paper_1.tex
Fixed references, converted to BibTex
	modified:   Paper_1.tex
	new file:   general.bib

\shorttitle{Effective Disk Alphas}
\shortauthors{Scott et al.}

@@

Using 3D hydrodynamic simulations of a cooling protostellar disk undergoing gravitational instabilities (GIs), we compute the effective Shakura and Sunyaev (1973) $\alpha$ due to gravitational stresses and compare them to predictions from an analytic local theory for thin disks by Gammie (2001).  Our goal is to determine how accurately a locally defined alpha can characterize mass and angular momentum transport by GIs in disks.  We examine disks with an imposed constant global cooling time (Mej\'{i}a et al. 2005).  Models were computed with grids of different spatial resolution to investigate how the computed alpha is affected by numerical resolution.

@@

While detailed three-dimensional hydrodynamic modeling of protostellar disks can
provide considerable insight into the physical and thermal states of these disks,
computing resources limit such studies to snapshots
spanning timescales much shorter than a disk's evolutionary lifetime.  In contrast, 
approaches that replace detailed physics with statistical parameters that
encapsulate this physics allow for calculations covering
much longer timeframes.  Both approaches can play important roles in broadening our 
understanding of these complex nonlinear systems, provided a connection between the detailed physics 
and statistical parameters can be demonstrated and understood.   In what follows, we
examine the applicability of an $\alpha$-disk formulation (Shakura \& Shaviv 1973) to  the
evolution of protostellar disks subject to gravitational instabilities.

Gravitational instabilities (GIs) play an important, and at times dominant,
role in driving the structural and thermal
evolution of  protostellar disks \citep[for a review, see][]{Dur10}.  
%In particular, torques arising from GIs provide a very effective means of redistributing matter and 
%angular momentum within these disks 
%\citep{Lod04, Lod05, Vor05, Vor06, Bol09}.  
Thermal processes play the primary role in regulating the
amplitude and outcome of these instabilities  (Pickett
et al. 1998, 2000a,b, 2003, see also Nelson et al. 1998, 2000). 
A link between a disk's thermodynamic state and its susceptibility to GIs
is provided by the Toomre $Q$-parameter (Toomre 1981);  $Q = c_s \kappa / \pi G \Sigma$,
where $c_s$ is the sound speed, $\kappa$ is the epicyclic frequency ($\sim$ the rotation frequency 
$\Omega$ in a nearly Keplerian disk), and $\Sigma$ is the disk surface mass density. 
For $Q \al$  1.5 to 1.7,  small density perturbations in a disk
grow exponentially on a time scale comparable to the rotation period.
These perturbations manifest themselves as multi-arm spirals with a predominantly trailing pattern that
transports angular momentum outward by
gravitational torques (e.g.,Larson 1984, Boss 1984, Durisen et al. 1986, Papaloizou \& Savonije
1991, Laughlin \& Bodenheimer 1994, Laughlin et al., 1998; Nelson
et al., 1998; Pickett et al., 1998, add refs?).
The long range nature of these torques cause nonlinear
spiral structures to develop over a broad range of radii (Laughlin et al. 1998, Nelson et al. 1998,
2000; Pickett et al. 1998, 2000, 2003).

Numerous studies
have shown that thermal physics can control GI amplitudes
by a {\it balance} of heating and cooling (e.g., Tomley et al. 1991, 1994;
Pickett et al., 1998, 2000,
2003; Gammie, 2001; Boss, 2002; Rice et al., 2003; Mej\'{i}a et al., 2005; Boley
et al., 2006, 2007a; Stamatellos and Whitworth, 2008; Cossins et al., 2009).
Heating by GIs tends to increase $c_s$, thus increasing $Q$. If disk heating persists,
the instability will shut off once $Q$ becomes sufficiently large.  However, if the
disk cools radiatively, then heating and cooling can reach a balance at nonlinear
wave amplitude. GI-activity can sustain this balance  at a relatively constant,
but unstable, value of $Q$ (e.g., Paczynski, 1978; Lin and Pringle; Pringle, 1981;
see also Goldreich and  Lynden-Bell 1965).  
For sufficiently low cooling rates, the disk can exist in a state of quasi-equilibrium, 
where cooling is 
balanced by heating arising from the inward flow of matter and shocks induced by the GIs
\citep{Gam01, Lod04, Ric05, Cos09, Vor10}.

Fully nonlinear hydrodynamic simulations of low-$Q$ disks show that multiple
modes can become unstable in the linear regime (e.g., Nelson et al.,
1998; Pickett et al., 1998; additional refs?).  Even if the initial growth is dominated by
one mode, numerous modes usually appear in the nonlinear regime. As a consequence, 
if a disk achieves a quasi-steady balance between heating and cooling, 
then there tends to be power at
all $m$-values resolved by a given numerical method (e.g., Mej\'{i}õa et al., 2005;
Boley et al., 2006).  
In this  {\it asymptotic state}, GI-active disks develop overlapping
density structures of different strengths, geometries and coherence,
producing
strongly nonlinear turbulence that pervades the entire disk
(e.g., Pickett et al. 2003, Mej\'{i}a et al. 2005, Boley et al. 2006).
Gammie (2001) refers to this state as {\it gravitoturbulence}.  
Gravitoturbulence provides a possible bridge between
the detailed physics of GIs and the viscous transport parameter $\alpha$.

Shakura \& Syunyaev (1973) first parameterized turbulent transport in a steady-state
disk with the  parameter $\alpha$, where the source of the
turbulence is unspecified.
Although their original work was directed towards X-ray bright
accretion disks around black holes, this heuristic
approach has proved fruitful in many 
astrophysical situations by allowing for
analytic solutions and relatively easy numerical modeling
over time-spans inaccessible to detailed modeling.
In the $\alpha$ disk formalism, when the accretion rate
$\dot{M}$ is constant in both time and radius, $\dot{M} = 
3\pi\alpha c_s^2 \Sigma / \Omega$ where $c_s$ is the sound speed, $\Sigma$ the 
surface density, and $\Omega$ the rotational frequency. 
The $\alpha$ 
formalism makes no assumption
about the nature of the angular momentum transport, only that {\it heating and cooling
are dominated by local processes}; the validity
of an $\alpha$-disk picture requires self-consistency of local energy dissipation.

Dissipative locality in GI-active disks has not been rigorously tested and 
remains an open question. 
Several studies support the idea that GI
transport in real disks is, in many important respects, an intrinsically global phenomenon and
cannot be properly treated by a local $\alpha$-like prescription (Laughlin \& Rozyczka 1996, Balbus
\& Papaloizou 1999, Lodato \& Rice 2005, Mej\'{i}a et al. 2005, Boley et al. 2006, Cai et al.
2007).  This almost certainly applies to the embedded phase of both high and low-mass
stars when infall from the collapsing protostellar cloud feeds mass into the disk at a rapid
rate (e.g., Yorke et al. 1993, Laughlin \& Bodenhiemer 1994, Yorke \& Bodenheimer 1999,
Vorobyov \& Basu 2005, 2006, Kratter \& Matzner 2007), and disks observed in the embedded
phase can be quite massive relative to their central stars (e.g., Osorio et al. 2003). 
However, despite the long-distance nature of gravitational torques, theoretical
and numerical results suggest that in protostellar disks where heating and cooling are in
balance it may be valid to represent GI-induced transport by an {\it effective viscosity}, a local process
(e.g., Gammie 2001; Rice et al. 2005; Lodato \& Rice 2004).
Indeed, Gammie (2001) derived an effective alpha for the case of a thin (effectively 2D)
gravitationally unstable disk where the heating due to local turbulent dissipation caused by
GIs was balanced by the local cooling rate.  Modifications to this derivation have since been
proposed by {\bf (Tom: include other derivations and refs)}.   Recent simulations (e.g., Cossins
et al. 2009; Vorobyov 2010) suggest that, at least in the case of imposed cooling,
$\alpha_{eff}$ roughly converges to the value predicted by Gammie.

@@

{\bf (Tom: add text concerning need for convergence approach)}
Here we report the results of a convergence study designed to determine
whether GI-active disks, chaotic by nature, achieve statistical equilibrium; if so, to what
level do these disks approximate the gravitoturbulence of Gammie (2001) and with what distribution
of $\alpha_{eff}$; and to characterize how the GI cascades from low-order to high-order modes. 
To this end, we follow the evolution of a GI-active disk, subject to constant cooling, in its asymptotic state,
where cooling and heating are essentially balanced.  Simulations are
performed using four different resolutions in the azimuthal direction.

@@

Simulations were conducted with the three-dimensional hydrodynamics 
code used by the
Indiana University Hydrodynamics Group in several previous studies
\citep{Pic98, Pic00, Mej04, Mej05, Bol06, Cai08}.  This code
uses a second-order, explicit, Eulerian scheme to solve Poisson's equation, 
an equation of state,
and the equations of hydrodynamics in conservative
form on a uniform cylindrical grid.  The code assumes mirror symmetry about the equatorial plane.
It includes both
self-gravity and artificial viscosity; the latter serves to 
mediate shocks.
Source and flux terms (Norman \&Winkler 1986) are computed separately in an
explicit, second-order time integration (van Albada et al. 1982; Christodoulou 1991; Yang 1992).

We assume an equation of state $P = (\gamma -1)\epsilon$, 
where $P$ is the pressure, $\epsilon$ is the internal energy density, and $\gamma$ 
is the ratio of specific heats, given here by $\gamma = 5/3$.  
The model disks are cooled by decreasing their internal energy according to 
the prescription $d\epsilon/dt = \epsilon / t_{cool}$, where $t_{cool}$ is the global cooling time in Outer Rotation Periods (orp).  For the grid resolutions used here, one orp is defined as the rotational period at radial zone 200.

@@

Because this is a polytropic disk with an idealized cooling time, only the ratio of disk to stellar mass, given here by 0.153, is specified.  In what follows, quantities specified with real physical units are derived assuming a 1 $M_\odot$ central star and a 0.153  $M_\odot$ disk.   In this case, the orp is defined at $r = 33$ AU and
one orp $\approx$ 303 years.  Models have $t_{cool} = 2$ orp.   {\bf (Scott: Please check these numbers)}
The disk has
an initial surface density $\Sigma \sim r^{ -1}$, and initial inner and outer radii of 2.3 and 40 AU, respectively.

Initial conditions for the disk's structure and thermodynamic state
were set using an equilibrium star plus disk model generated using a modified Haschisu (1986)
self-consistent field (SCF) relaxation method (Pickett et al. 1996, 2003;
Mej\'{i}a (2004);  Mej\'{i}a et al. (2005); Cai 2006).  
Here  the density and angular
momentum distributions are iteratively
solved, using the specified equation of state,  until convergence is achieved. 
This  equilibrium disk was then seeded with small random perturbations to
allow GIs to grow.  The resultant marginally unstable disk served as the $t = 0$ load for the hydro code.

The initially unstable disk model passed through several phases of evolution
(see also Pickett et al., 2003; Mej\'{i}a et al., 2005).  The initial load is unstable to
the growth of axisymmetric structures. During this axisymmetric phase, 
which lasts several orps,
the disk shrinks slightly, in an axisymmetric manner, until it becomes unstable
to nonaxisymmetric perturbations.    When this condition is met, the disk undergoes a strong burst 
of GIs that predominantly manifest themselves in
one or two discrete global spiral modes.  The disk expands violently during the burst phase,
producing a significant rearrangement of the disk's mass distribution on a timescale of one or
two orps. 
This is then followed by a period of several orps where heating temporarily washes
out some of the nonaxisymmetry in the disk.  Finally the disk settles into
a quasi-steady, long-lived asymptotic state of sustained GI-activity
over a large part of the disk, with an overall balance of heating and cooling.

Simulations were run with four different angular resolutions, with grid sizes in $(r,\phi,z)$
of  $(j_{max}, k_{max}, l_{max}) =$
 (512,64,64), (512,128,64), (512,256,64), and (512,512,64).
Because of the large computational overhead associated with starting each simulation from
$t = 0$,
the $l_{max} =$ 128 simulation was run through the axisymmetric, burst and transition
phases. Simulations using $l_{max} =$ 64, 256, and 512 
all begin by interpolating the $l_{max} =$ 128 simulation
to higher or lower azimuthal resolution at 9.6 ORPs, near the end of the transition
phase. Although this strategy saves a large amount of computational time, it limits
the scope of the analysis to the asymptotic phase and
it introduces a perturbation to the disk which takes several
orps to subside. By 12 orps, the perturbation has dissipated and all four disks
have transitioned to the asymptotic state.  All simulations were 
run through $\sim 18$ orps.

Figure 1 shows $Q$, as a function of radius, at $t \sim$ 18 orp, while
Table 1 lists $Q_{avg}$, the value of $Q$ of for the asymptotic disk, time-averaged of
12 -- 19 orp and spatially averaged over 10 -- 40 AU, as a function of radius
for each resolution, 
{\bf (Scott: Figure 1 needs to be redone - I have extracted the 
plot from your chapter 3 and it only shows the 128 model.  We should show all four resolutions
or ditch this figure)}.

@@

Figure 2
%\ref{fig:DensityPlots}, 
shows midplane and radial densities  at $t \sim 18$ orps  
for the $l_{max} =$ 64, 128, 256, and 512 simulations. 
The lower resolution (smaller $l_{max}$) simulations
display longer wavelength spiral structures, with larger amplitudes and more coherence than those
seen in higher resolution 
simulations.  
Higher angular resolution
allows GI power to spread to higher-order azimuthal modes, as demonstrated 
by a Fourier decomposition of the density.
The strengths of these modes are given by their global Fourier amplitudes,
\begin{equation}
A_m = \frac{(a_m^2 + b_m^2)^{1/2}}{\pi\int\rho_o \varpi {\rm d}\varpi {\rm d}z},
\end{equation}
where
\renewcommand{\theequation}{\arabic{equation}\alph{subequation}}
\setcounter{subequation}{2}
\begin{equation}
a_m = \int \rho \cos(m\phi)\varpi {\rm d}\varpi {\rm d}z {\rm d}\phi,
\end{equation}
\addtocounter{equation}{-1}
\addtocounter{subequation}{1}
\begin{equation}
b_m =\int \rho \sin(m\phi)\varpi {\rm d}\varpi {\rm d}z {\rm d}\phi.
\end{equation}
\renewcommand{\theequation}{\arabic{equation}}
Here  $\rho_o$ is the axisymmetric (zero-order)  component of the density.
In what follows, $A_m$ is averaged over time to 
suppress fluctuations on the dynamic time scale due
to the chaotic nature of GIs in the asymptotic  phase.
In particular, $\langle A_{tot} \rangle$ will represent 
the power over the radial range 10 -- 40 AU,
integrated from $ m = 2$ to $l_{max}/2$ and
averaged over the time range 12 -- 19 orps.  
The density distribution cannot be resolved in azimuth
on angular scales less than $2\pi/l_{max}$, therefore $A_m$ is
limited to modes $m \le l_{max}/2$.  
Because the central star is artificially fixed to the grid center, the $m = 1$ mode 
cannot be be realistically treated; it is excluded from
 $\langle A_{tot} \rangle$.

@@

Previous studies have shown that disks in the asymptotic state have power at
all resolvable $m$-values  (e.g., Mej\'{i}õa et al., 2005;
Boley et al., 2006).  Table 1 lists $\langle A_{tot} \rangle$ along with 
$\langle A_{2-7} \rangle  /  \langle A_{tot} \rangle$,  the fraction of the total
power falling into low-order ($m = 2$ -- 7) modes, for each resolution. 
While the total power is approximately independent of
resolution, $\langle A_{tot} \rangle = 1.90 \pm 0.07$, 
the modal distribution of this power shows marked differences.
In particular, the fractional power in lower order modes decreases monotonically
with increasing resolution.  Over three quarters of $\langle A_{tot} \rangle$ resides
in modes $m =$ 2 -- 7 for the $l_{max} = 64$ simulation, while approximately
half resides in these modes when $l_{max} =$ 512.

The shift of power from lower to higher order modes as the resolution of
the computational grid is increased arises from the fact that, while
$\langle A_{tot} \rangle$ is approximately independent of $l_{max}$, 
the number of degrees of freedom available for this power, $l_{max}/2$, 
increases linearly with the resolution.  
Thus power is naturally spread from lower order to higher order modes
as the available $m$-values increases.  
Indeed, even if the disk is dominated by one or two low-order modes at the
time of outburst, this power cascades to higher-order modes (e.g., 
Laughlin et al. 1997, 1998; see also Laughlin and Korchagin 1996).

Given the resolution dependence of how power is
distributed,  we must examine 
what resolution is sufficient to properly capture the evolution of these disks.
Table 1 shows that while the fractional power in lower order modes
decreases with each doubling of $l_{max}$, the amount of change in going from $l_{max} = 256$ to 512 
is significantly less than the change going from $l_{max} = 64$ to 128,
with a strong suggestion
that this diminution will not proceed much further with higher azimuthal resolution.  
Indeed, a  Richardson extrapolation {\bf (add reference)} 
of the tabulated values of  $\langle A_{2-7} \rangle / \langle A_{tot} \rangle$
suggests a limiting value of  $\langle A_{2-7} \rangle / \langle A_{tot} \rangle \sim$ 0.43  
for increasingly higher resolutions.
This is displayed graphically in Figure 4, where the
fractional power in lower order modes is plotted against the  normalized grid 
spacing for each value of $l_{max}$.
The grid spacing  is normalized such that spacing of the $l_{max}$ = 512 grid
equals unity.  Convergence occurs as the normalized spacing goes to zero.
The Richardson Extrapolation value  of $\langle A_{2-7}\rangle  / \langle A_{tot}\rangle  = 0.43$ is
plotted at a normalized grid spacing  equal to zero.   More than half of the modal power in
a converged disk resides in higher-order modes, modes 
that help define a state of gravitoturbulence.
The fact that
the $l_{max} =$ 512 simulation has
$\langle A_{2-7} \rangle / \langle A_{tot} \rangle$ near the convergence value
lends some confidence in the use of the $l_{max} =$ 512 model in
examining the convergence of $\alpha$.

@@

The gravitational torque from equation \ref{torque1} can be used to determine the
gravitational stress by
\begin{equation}
\begin{split}
 T_{\varpi\phi}^{grav}\left( \varpi \right) & = \frac{1}{2\pi\varpi^2}C_Z\\
                                & = - \frac{1}{2\pi\varpi^2} \int \rho\frac{\partial\Phi}{\partial\phi} {\rm d}V.
\end{split}
\end{equation}
{\bf (There is a problem with this equation)}
Figure 6 shows the effective $\alpha$s for the four different grid resolutions, time
averaged from 12 to 19 orps.  
For comparison, the predicted values by Gammie (2001)
for the strongly self-gravitating limit (lower curve) and the non-self-gravitating limit
(upper curve) are shown. Note that predictions from are based upon a thin
disk approximation, where the local balance of heating and cooling dominates the
energetics.

,

>>>>

\shorttitle{Effective Disk Alphas}
\shortauthors{Michael et al.}

@@

Using 3D hydrodynamic simulations of a cooling protostellar disk undergoing gravitational instabilities (GIs), we compute the effective \citet{shakura1973} $\alpha$ due to gravitational stresses and compare them to predictions from an analytic local theory for thin disks by \citet{gammie2001}.  Our goal is to determine how accurately a locally defined alpha can characterize mass and angular momentum transport by GIs in disks.  We examine disks with an imposed constant global cooling time \citep{mejia2005}.  Models were computed with grids of different spatial resolution to investigate how the computed alpha is affected by numerical resolution.

@@

While detailed three-dimensional hydrodynamic modeling of protostellar disks can provide considerable insight into the physical and thermal states of these disks, computing resources limit such studies to snapshots spanning timescales much shorter than a disk's evolutionary lifetime.  In contrast, approaches that replace detailed physics with statistical parameters that encapsulate this physics allow for calculations covering much longer timeframes.  Both approaches can play important roles in broadening our understanding of these complex nonlinear systems, provided a connection between the detailed physics and statistical parameters can be demonstrated and understood. In what follows, we examine the applicability of an $\alpha$-disk formulation \citep{shakura1973} to  the
evolution of protostellar disks subject to gravitational instabilities.

Gravitational instabilities (GIs) play an important, and at times dominant, role in driving the structural and thermal
evolution of  protostellar disks \citep[for a review, see][]{durisen2005}.
%In particular, torques arising from GIs provide a very effective means of redistributing matter and 
%angular momentum within these disks 
%\citep{Lod04, Lod05, Vor05, Vor06, Bol09}.  
Thermal processes play the primary role in regulating the amplitude and outcome of these instabilities \citep{pickett1998,pickett2000,mejia2005,nelson1998,nelson2000a}. A link between a disk's thermodynamic state and its susceptibility to GIs is provided by the Toomre $Q$-parameter \citep{toomre1981};  $Q = c_s \kappa / \pi G \Sigma$, where $c_s$ is the sound speed, $\kappa$ is the epicyclic frequency ($\sim$ the rotation frequency 
$\Omega$ in a nearly Keplerian disk), and $\Sigma$ is the disk surface mass density. For $Q \al$  1.5 to 1.7,  small density perturbations in a disk grow exponentially on a time scale comparable to the rotation period. These perturbations manifest themselves as multi-arm spirals with a predominantly trailing pattern that transports angular momentum outward by gravitational torques \citep[e.g.]{larson1984,boss1984,durisen1986,papaloizou1991, laughlin1994,nelson1998,pickett1998}. The long range nature of these torques cause nonlinear
spiral structures to develop over a broad range of radii \citep{laughlin1998,nelson1998,nelson2000a,pickett1998, pickett2000,pickett2003}.

Numerous studies have shown that thermal physics can control GI amplitudes by a {\it balance} of heating and cooling \citep[for example]{tomley1991,tomley1994,pickett1998,pickett2000, pickett2003,gammie2001,boss2002,rice2003b,mejia2005,boley2006,boley2007, stamatellos2008,cossins2009}.
Heating by GIs tends to increase $c_s$, thus increasing $Q$. If disk heating persists, the instability will shut off once $Q$ becomes sufficiently large.  However, if the disk cools radiatively, then heating and cooling can reach a balance at nonlinear wave amplitude. GI-activity can sustain this balance  at a relatively constant,
but unstable, value of $Q$ \citep{paczynski1978, lin1981,goldreich1965}.  
For sufficiently low cooling rates, the disk can exist in a state of quasi-equilibrium, where cooling is 
balanced by heating arising from the inward flow of matter and shocks induced by the GIs
\citep{gammie2001, lodato2004, rice2005, cossins2009, vorobyov2010}.

Fully nonlinear hydrodynamic simulations of low-$Q$ disks show that multiple
modes can become unstable in the linear regime \citep[e.g.]{nelson1998, pickett1998} {\bf additional refs?}.  Even if the initial growth is dominated by one mode, numerous modes usually appear in the nonlinear regime. As a consequence, if a disk achieves a quasi-steady balance between heating and cooling, then there tends to be power at all $m$-values resolved by a given numerical method \citep{mejia2005,boley2006}.  
In this  {\it asymptotic state}, GI-active disks develop overlapping density structures of different strengths, geometries and coherence,producing strongly nonlinear turbulence that pervades the entire disk \citep[e.g.]{pickett2003, mejia2005,boley2006}.
\citet{gammie2001} refers to this state as {\it gravitoturbulence}. Gravitoturbulence provides a possible bridge between the detailed physics of GIs and the viscous transport parameter $\alpha$.

\citet{shakura1973} first parameterized turbulent transport in a steady-state disk with the  parameter $\alpha$, where the source of the turbulence is unspecified. Although their original work was directed towards X-ray bright
accretion disks around black holes, this heuristic approach has proved fruitful in many astrophysical situations by allowing for analytic solutions and relatively easy numerical modeling over time-spans inaccessible to detailed modeling. In the $\alpha$ disk formalism, when the accretion rate $\dot{M}$ is constant in both time and radius, $\dot{M} = 3\pi\alpha c_s^2 \Sigma / \Omega$ where $c_s$ is the sound speed, $\Sigma$ the surface density, and $\Omega$ the rotational frequency. The $\alpha$ formalism makes no assumption about the nature of the angular momentum transport, only that {\it heating and cooling are dominated by local processes}; the validity of an $\alpha$-disk picture requires self-consistency of local energy dissipation.

Dissipative locality in GI-active disks has not been rigorously tested and remains an open question. 
Several studies support the idea that GI transport in real disks is, in many important respects, an intrinsically global phenomenon and cannot be properly treated by a local $\alpha$-like prescription \citep{laughlin1996, balbus1999, lodato2005, mejia2005, boley2006, cai2008}.  This almost certainly applies to the embedded phase of both high and low-mass stars when infall from the collapsing protostellar cloud feeds mass into the disk at a rapid
rate \citep[e.g.]{yorke1993, laughlin1994, yorke1999,vorobyov2005,vorobyov2006}, and disks observed in the embedded phase can be quite massive relative to their central stars \citep[e.g.]{osorio2003}. 
However, despite the long-distance nature of gravitational torques, theoretical and numerical results suggest that in protostellar disks where heating and cooling are in balance it may be valid to represent GI-induced transport by an {\it effective viscosity}, a local process \citep[e.g.]{gammie2001, rice2005, lodato2004}. Indeed, \citet{gammie2001} derived an effective alpha for the case of a thin 2D gravitationally unstable disk where the heating due to local turbulent dissipation caused by GIs was balanced by the local cooling rate.  Modifications to this derivation have since been proposed by {\bf (Tom: include other derivations and refs)}.   Recent simulations \citep[e.g.]{cossins2009, vorobyov2010} suggest that, at least in the case of imposed cooling, $\alpha_{eff}$ roughly converges to the value predicted by \citeauthor{gammie2001}.

@@

{\bf (Tom: add text concerning need for convergence approach)}
Here we report the results of a convergence study designed to determine whether GI-active disks, chaotic by nature, achieve statistical equilibrium; if so, to what level do these disks approximate the gravitoturbulence of \citet{gammie2001} and with what distribution of $\alpha_{eff}$; and to characterize how the GI cascades from low-order to high-order modes. To this end, we follow the evolution of a GI-active disk, subject to constant cooling, in its asymptotic state, where cooling and heating are essentially balanced.  Simulations are performed using four different resolutions in the azimuthal direction.

@@

Simulations were conducted with the three-dimensional hydrodynamics code used by the Indiana University Hydrodynamics Group in several previous studies \citep{pickett1998, pickett2000, mejiaphd2004, mejia2005, boley2006, cai2008}.  This code uses a second-order, explicit, Eulerian scheme to solve Poisson's equation, an equation of state, and the equations of hydrodynamics in conservative form on a uniform cylindrical grid.  The code assumes mirror symmetry about the equatorial plane. It includes both self-gravity and artificial viscosity; the latter serves to 
mediate shocks. Source and flux terms \citep{norman1986} are computed separately in an explicit, second-order time integration \citep{albada1982,christodoulou1991,yangphd1992}.

We assume an equation of state $P = (\gamma -1)\epsilon$, where $P$ is the pressure, $\epsilon$ is the internal energy density, and $\gamma$ is the ratio of specific heats, given here by $\gamma = 5/3$.  The model disks are cooled by decreasing their internal energy according to the prescription $d\epsilon/dt = \epsilon / t_{cool}$, where $t_{cool}$ is the global cooling time in Outer Rotation Periods (ORP).  For the grid resolutions used here, one ORP is defined as the rotational period at radial zone 200.

@@

Because this is a polytropic disk with an idealized cooling time, only the ratio of disk to stellar mass, given here by 0.153, is specified.  In what follows, quantities specified with real physical units are derived assuming a 1 $M_\odot$ central star and a 0.153  $M_\odot$ disk. In this case, the ORP is defined at $r = 33$ AU and one ORP $\approx$ 303 years.  Models have $t_{cool} = 2$ ORP.   {\bf (Scott: Please check these numbers)}
The disk has an initial surface density $\Sigma \sim r^{ -1}$, and initial inner and outer radii of 2.3 and 40 AU, respectively.

Initial conditions for the disk's structure and thermodynamic state were set using an equilibrium star plus disk model generated using a modified \citet{hachisu1986} self-consistent field (SCF) relaxation method \citep{pickett1996,pickett2003,mejiaphd2004,mejia2005,cai2006}. Here  the density and angular momentum distributions are iteratively solved, using the specified equation of state,  until convergence is achieved. This  equilibrium disk was then seeded with small random perturbations to allow GIs to grow.  The resultant marginally unstable disk served as the $t = 0$ load for the hydro code.

The initially unstable disk model passed through several phases of evolution \citep[see also]{pickett2003, mejia2005}.  The initial load is unstable to the growth of axisymmetric structures. During this axisymmetric phase, 
which lasts several ORPs, the disk shrinks slightly, in an axisymmetric manner, until it becomes unstable to nonaxisymmetric perturbations.    When this condition is met, the disk undergoes a strong burst of GIs that predominantly manifest themselves in one or two discrete global spiral modes.  The disk expands violently during the burst phase, producing a significant rearrangement of the disk's mass distribution on a timescale of one or
two ORPs. This is then followed by a period of several ORPs where heating temporarily washes out some of the nonaxisymmetry in the disk.  Finally the disk settles into a quasi-steady, long-lived asymptotic state of sustained GI-activity over a large part of the disk, with an overall balance of heating and cooling.

Simulations were run with four different angular resolutions, with grid sizes in $(r,\phi,z)$ of  $(j_{max}, k_{max}, l_{max}) =$ (512,64,64), (512,128,64), (512,256,64), and (512,512,64). Because of the large computational overhead associated with starting each simulation from$t = 0$, the $l_{max} =$ 128 simulation was run through the axisymmetric, burst and transition phases. Simulations using $l_{max} =$ 64, 256, and 512 all begin by interpolating the $l_{max} =$ 128 simulation to higher or lower azimuthal resolution at 9.6 ORPs, near the end of the transition
phase. Although this strategy saves a large amount of computational time, it limits the scope of the analysis to the asymptotic phase and it introduces a perturbation to the disk which takes several ORPs to subside. By 12 ORPs, the perturbation has dissipated and all four disks have transitioned to the asymptotic state.  All simulations were 
run through $\sim 18$ ORPs.

Figure \ref{Final_Q} shows $Q$, as a function of radius, at $t \sim$ 18 ORP, while table \ref{tbl:ams} lists $Q_{avg}$, the value of $Q$ of for the asymptotic disk, time-averaged of 12 -- 19 ORP and spatially averaged over 10 -- 40 AU, as a function of radius for each resolution, 
{\bf (Scott: Figure 1 needs to be redone - I have extracted the 
plot from your chapter 3 and it only shows the 128 model.  We should show all four resolutions
or ditch this figure)}.

@@

Figure \ref{fig:DensityPlots}, shows midplane and radial densities  at $t \sim 18$ ORPs for the $l_{max} =$ 64, 128, 256, and 512 simulations. The lower resolution (smaller $l_{max}$) simulations display longer wavelength spiral structures, with larger amplitudes and more coherence than those seen in higher resolution simulations. Higher angular resolution allows GI power to spread to higher-order azimuthal modes, as demonstrated by a Fourier decomposition of the density. The strengths of these modes are given by their global Fourier amplitudes,

\begin{equation}
A_m = \frac{(a_m^2 + b_m^2)^{1/2}}{\pi\int\rho_o \varpi {\rm d}\varpi {\rm d}z},
\end{equation}
where
\renewcommand{\theequation}{\arabic{equation}\alph{subequation}}
\setcounter{subequation}{2}
\begin{equation}
a_m = \int \rho \cos(m\phi)\varpi {\rm d}\varpi {\rm d}z {\rm d}\phi,
\end{equation}
\addtocounter{equation}{-1}
\addtocounter{subequation}{1}
\begin{equation}
b_m =\int \rho \sin(m\phi)\varpi {\rm d}\varpi {\rm d}z {\rm d}\phi.
\end{equation}
\renewcommand{\theequation}{\arabic{equation}}

Here  $\rho_o$ is the axisymmetric (zero-order)  component of the density. In what follows, $A_m$ is averaged over time to suppress fluctuations on the dynamic time scale due to the chaotic nature of GIs in the asymptotic  phase.
In particular, $\langle A_{tot} \rangle$ will represent the power over the radial range 10 -- 40 AU, integrated from $ m = 2$ to $l_{max}/2$ and averaged over the time range 12 -- 19 ORPs. The density distribution cannot be resolved in azimuth on angular scales less than $2\pi/l_{max}$, therefore $A_m$ is limited to modes $m \le l_{max}/2$.  
Because the central star is artificially fixed to the grid center, the $m = 1$ mode cannot be be realistically treated; it is excluded from $\langle A_{tot} \rangle$.

@@

Previous studies have shown that disks in the asymptotic state have power at all resolvable $m$-values  \citep[e.g.]{mejia2005,boley2006}.  Table \ref{tbl:ams} lists $\langle A_{tot} \rangle$ along with $\langle A_{2-7} \rangle  /  \langle A_{tot} \rangle$,  the fraction of the total power falling into low-order ($m = 2$ -- 7) modes, for each resolution. While the total power is approximately independent of resolution, $\langle A_{tot} \rangle = 1.90 \pm 0.07$, 
the modal distribution of this power shows marked differences. In particular, the fractional power in lower order modes decreases monotonically with increasing resolution.  Over three quarters of $\langle A_{tot} \rangle$ resides
in modes $m =$ 2 -- 7 for the $l_{max} = 64$ simulation, while approximately half resides in these modes when $l_{max} =$ 512.

The shift of power from lower to higher order modes as the resolution of the computational grid is increased arises from the fact that, while $\langle A_{tot} \rangle$ is approximately independent of $l_{max}$, the number of degrees of freedom available for this power, $l_{max}/2$, increases linearly with the resolution. Thus power is naturally spread from lower order to higher order modes as the available $m$-values increases. Indeed, even if the disk is dominated by one or two low-order modes at the time of outburst, this power cascades to higher-order modes \citep[for example]{laughlin1997, laughlin1998,laughlin1996}.

Given the resolution dependence of how power is distributed,  we must examine what resolution is sufficient to properly capture the evolution of these disks. Table \ref{tbl:ams} shows that while the fractional power in lower order modes decreases with each doubling of $l_{max}$, the amount of change in going from $l_{max} = 256$ to 512 
is significantly less than the change going from $l_{max} = 64$ to 128, with a strong suggestion that this diminution will not proceed much further with higher azimuthal resolution. Indeed, a  Richardson extrapolation {\bf (add reference)} of the tabulated values of  $\langle A_{2-7} \rangle / \langle A_{tot} \rangle$ suggests a limiting value of  $\langle A_{2-7} \rangle / \langle A_{tot} \rangle \sim$ 0.43 for increasingly higher resolutions. This is displayed graphically in Figure 4, where the fractional power in lower order modes is plotted against the  normalized grid 
spacing for each value of $l_{max}$. The grid spacing  is normalized such that spacing of the $l_{max}$ = 512 grid
equals unity.  Convergence occurs as the normalized spacing goes to zero. The Richardson extrapolation value  of $\langle A_{2-7}\rangle  / \langle A_{tot}\rangle  = 0.43$ is plotted at a normalized grid spacing  equal to zero. More than half of the modal power in a converged disk resides in higher-order modes, modes that help define a state of gravitoturbulence. The fact that the $l_{max} =$ 512 simulation has $\langle A_{2-7} \rangle / \langle A_{tot} \rangle$ near the convergence value lends some confidence in the use of the $l_{max} =$ 512 model in examining the convergence of $\alpha$.

@@

The gravitational torque from equation \ref{torque1} can be used to determine the
gravitational stress by
\begin{equation}
\begin{split}
 T_{\varpi\phi}^{grav}\left( \varpi \right) & = \frac{1}{2\pi\varpi^2}C_Z\\
                                & = - \frac{1}{2\pi\varpi^2} \int \rho\frac{\partial\Phi}{\partial\phi} {\rm d}V.
\end{split}
\end{equation}
{\bf (There is a problem with this equation)}
Figure 6 shows the effective $\alpha$s for the four different grid resolutions, time averaged from 12 to 19 ORPs. For comparison, the predicted values by \citet{gammie2001} for the strongly self-gravitating limit (lower curve) and the non-self-gravitating limit (upper curve) are shown. Note that predictions from are based upon a thin
disk approximation, where the local balance of heating and cooling dominates the energetics.
